---
title: "MAD Taxonomic Analysis — Pain: Fecal vs Cecal"
author: "Wes Tom"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6)
data_dir <- Sys.getenv("MAD_DATA_DIR", unset = "")
if (!nzchar(data_dir)) data_dir <- normalizePath(".")
output_root <- Sys.getenv("MAD_OUTPUT_DIR", unset = "")
if (!nzchar(output_root)) output_root <- data_dir
data_dir <- normalizePath(data_dir)
dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
output_root <- normalizePath(output_root)
knitr::opts_knit$set(root.dir = data_dir)
```

# Overview

SingleM relative-abundance profiles (`otu_table.csv`) linked to `tax_table.csv` and `subject_data.csv`.

**Cohort:** pain subjects with paired fecal and cecal samples (`fvc_ps`, `pop == "pain"`).

**Primary comparison:** GI site (fecal vs cecal). Taxa are relative abundances; MaAsLin2 uses `normalization = "NONE"`, `transform = "NONE"`.

**Direction:** MaAsLin2 reference = `sample_type,Fecal`; paired Wilcoxon `median_diff` = cecal − fecal.

Outputs: `analysis_output/` (MaAsLin2 in `maaslin2/`, consensus in `consensus/`, RF in `random_forest/`).

---

```{r packages}
pkgs <- c(
  "phyloseq", "microViz", "ampvis2", "dplyr", "tidyr", "tibble", "purrr",
  "ggplot2", "ggpubr", "vegan", "Maaslin2", "randomForest", "ComplexHeatmap"
)
missing <- pkgs[!vapply(pkgs, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing)) stop("Install missing packages: ", paste(missing, collapse = ", "))
library(phyloseq)
library(microViz)
library(ampvis2)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(ggpubr)
library(vegan)
library(Maaslin2)
library(randomForest)
source(file.path(data_dir, "R", "microbiome_helpers.R"))
```

```{r paths}
output_dir  <- file.path(output_root, "analysis_output")
maaslin_dir <- file.path(output_dir, "maaslin2")
rf_dir      <- file.path(output_dir, "random_forest")
consensus_dir <- file.path(output_dir, "consensus")
dir.create(maaslin_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(rf_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(consensus_dir, recursive = TRUE, showWarnings = FALSE)
setwd(data_dir)
```

# 1. Load data

```{r load-data}
mad_objs <- load_mad_phyloseq(data_dir)
mad_ps <- mad_objs$mad_ps
fvc_ps <- mad_objs$fvc_ps
meta_priority <- mad_objs$meta_priority
fvc_ps
```

```{r cohort-summary}
tibble(
  cohort = "Pain fecal+cecal",
  n_samples = nsamples(fvc_ps),
  n_subjects = length(unique(sample_data(fvc_ps)$ssn)),
  n_fecal = sum(sample_data(fvc_ps)$sample_type == "Fecal"),
  n_cecal = sum(sample_data(fvc_ps)$sample_type == "Cecal")
) %>% knitr::kable()
```

# 2. Alpha diversity

```{r alpha-diversity}
fvc_amp <- amp_load(fvc_ps)
fvc_alpha <- amp_alphadiv(fvc_amp, measure = c("uniqueotus", "shannon"))

p_u <- ggviolin(fvc_alpha, x = "sample_type", y = "uniqueOTUs",
  fill = "sample_type", palette = c("gold", "darkred"), add = "boxplot") +
  stat_compare_means(method = "wilcox.test") +
  ggtitle("Richness — fecal vs cecal")
p_s <- ggviolin(fvc_alpha, x = "sample_type", y = "Shannon",
  fill = "sample_type", palette = c("gold", "darkred"), add = "boxplot") +
  stat_compare_means(method = "wilcox.test") +
  ggtitle("Shannon — fecal vs cecal")

ggarrange(p_u, p_s, ncol = 2)
ggsave(file.path(output_dir, "alpha_diversity_pain_gi.pdf"), width = 10, height = 5)
```

# 3. Beta diversity

Aitchison distances on CLR-transformed species abundances (microViz).

```{r permanova}
perm_fvc <- run_permanova_microviz(
  fvc_ps,
  c("sample_type", "calprotectin_high", "phq_total", "gad_total"),
  "Pain fecal vs cecal"
)
perm_fvc$table %>%
  mutate(cohort = "pain_fvc") %>%
  write.csv(file.path(output_dir, "permanova_pain_fecal_vs_cecal.csv"), row.names = FALSE)
knitr::kable(perm_fvc$table, caption = "PERMANOVA — pain cohort")
```

```{r pca}
fvc_ps %>%
  tax_agg("species") %>%
  tax_transform("clr") %>%
  ord_calc() %>%
  ord_plot(plot_taxa = 1:5, colour = "sample_type", size = 1.5) +
  stat_ellipse(aes(colour = sample_type), linewidth = 0.3) +
  ggtitle("Pain cohort — CLR PCA (fecal vs cecal)")
ggsave(file.path(output_dir, "pca_pain_fecal_cecal.pdf"), width = 8, height = 6)
```

# 4. Composition heatmap

```{r heatmap}
htmp <- fvc_ps %>%
  comp_heatmap(
    taxa = tax_top(fvc_ps, n = 15),
    grid_col = NA,
    colors = heat_palette(palette = viridisLite::turbo(11)),
    row_names_gp = grid::gpar(fontsize = 10),
    sample_anno = sampleAnnotation(
      GI_location = anno_sample_cat("sample_type", col = c(Fecal = "gold", Cecal = "darkred")),
      Calprotectin = anno_sample_cat("calprotectin_high", col = c(`1` = "salmon", `0` = "forestgreen"))
    )
  )
png(file.path(output_dir, "heatmap_pain_top15.png"), width = 14, height = 10, units = "in", res = 300)
ComplexHeatmap::draw(htmp, merge_legends = TRUE)
dev.off()
```

# 5. Differential abundance

| Method | Role |
|--------|------|
| MaAsLin2 (unpaired) | Multivariable LM + covariates |
| MaAsLin2 paired (`ssn` random effect) | Within-subject GI site |
| Paired Wilcoxon | Within-subject screen |
| LEfSe-style | Kruskal–Wallis + BH |
| Random forest | Supporting importance |

Consensus: **≥2 methods** at *q* < 0.25 for `sample_type`.

## 5a. MaAsLin2

```{r maaslin-fvc}
maaslin_fvc <- run_maaslin(
  ps = fvc_ps,
  fixed_effects = c(
    "sample_type", "calprotectin", "calprotectin_high",
    "age", "sex", "bmi", "obese",
    "phq_total", "phq_difficulty", "gad_total", "gad_difficulty", "household_shared"
  ),
  output_subdir = "pain_fecal_vs_cecal",
  maaslin_dir = maaslin_dir,
  reference = "sample_type,Fecal",
  random_effects = "household_id"
)
plot_maaslin_volcano(
  maaslin_fvc, "MaAsLin2 — Pain (fecal vs cecal)",
  file.path(output_dir, "maaslin_volcano_pain_fvc.pdf"),
  meta_priority
)
```

```{r maaslin-fvc-paired}
out_paired <- file.path(maaslin_dir, "pain_fecal_vs_cecal_paired_ssn")
dir.create(out_paired, recursive = TRUE, showWarnings = FALSE)
dat <- as.data.frame(otu_table(fvc_ps))
if (ncol(dat) == nsamples(fvc_ps)) dat <- t(dat)
meta_p <- prepare_maaslin_metadata(
  fvc_ps,
  fixed_effects = c("sample_type", "calprotectin", "calprotectin_high",
                    "age", "sex", "bmi", "obese", "phq_total", "gad_total")
)
meta_p <- meta_p[rownames(dat), , drop = FALSE]
maaslin_fvc_paired <- Maaslin2(
  input_data = dat,
  input_metadata = meta_p,
  output = out_paired,
  min_abundance = 0, min_prevalence = 0.1, max_significance = 0.25,
  normalization = "NONE", transform = "NONE",
  analysis_method = "LM",
  fixed_effects = c("sample_type"),
  random_effects = c("ssn"),
  reference = "sample_type,Fecal",
  plot_scatter = FALSE
)
```

## 5b. Paired Wilcoxon and LEfSe-style

```{r diffabund-pain}
paired_fvc <- paired_wilcox_screen(fvc_ps, "sample_type", p_thresh = 0.25)
lefse_fvc  <- lefse_screen(fvc_ps, "sample_type", p_thresh = 0.25)
write.csv(paired_fvc, file.path(output_dir, "paired_wilcox_pain_gi.csv"), row.names = FALSE)
write.csv(lefse_fvc,  file.path(output_dir, "lefse_style_pain_gi.csv"), row.names = FALSE)
```

# 6. Random forest

```{r random-forest}
rf_fvc <- run_rf_classifier(fvc_ps, "sample_type", "pain_fecal_vs_cecal")
if (!is.null(rf_fvc)) {
  plot_rf_importance(rf_fvc, out_file = file.path(rf_dir, "rf_importance_pain_gi_site.pdf"))
  write.csv(rf_fvc$importance, file.path(rf_dir, "rf_importance_pain_gi_site.csv"), row.names = FALSE)
}
```

# 7. Consensus

```{r consensus}
q_thresh <- 0.25
pain_hits <- list(
  maaslin2 = read_maaslin_hits(maaslin_dir, "pain_fecal_vs_cecal", "sample_type", q_thresh) %>%
    distinct(feature, .keep_all = TRUE),
  maaslin2_paired = read_maaslin_hits(maaslin_dir, "pain_fecal_vs_cecal_paired_ssn", "sample_type", q_thresh) %>%
    distinct(feature, .keep_all = TRUE),
  paired_wilcox = paired_fvc %>% distinct(feature, .keep_all = TRUE),
  lefse_style = lefse_fvc %>% distinct(feature, .keep_all = TRUE)
)
pain_hits <- pain_hits[vapply(pain_hits, nrow, integer(1)) > 0]

consensus_pain <- build_consensus(
  cohort_label = "pain_fecal_vs_cecal",
  comparison = "sample_type",
  method_hits = pain_hits,
  rf_imp = if (!is.null(rf_fvc)) rf_fvc$importance else NULL,
  min_methods = 2
)
write.csv(consensus_pain, file.path(consensus_dir, "consensus_pain_gi_site.csv"), row.names = FALSE)
write.csv(
  consensus_pain %>% filter(n_methods >= 3),
  file.path(consensus_dir, "consensus_pain_gi_site_strong.csv"),
  row.names = FALSE
)
write.csv(method_overlap(pain_hits), file.path(consensus_dir, "method_overlap_pain.csv"), row.names = FALSE)

if (nrow(consensus_pain) > 0) {
  ggplot(consensus_pain %>% slice_head(n = 20),
         aes(x = reorder(feature, n_methods), y = n_methods, fill = rf_top_support)) +
    geom_col() + coord_flip() +
    labs(title = "Consensus taxa — pain fecal vs cecal",
         subtitle = ">=2 methods (MaAsLin2, paired Wilcoxon, LEfSe-style)",
         x = NULL, y = "# methods") +
    theme_bw()
  ggsave(file.path(consensus_dir, "consensus_pain_top20.pdf"), width = 9, height = 7)
  knitr::kable(consensus_pain %>% slice_head(n = 25))
}
```

**Output files**

- `consensus/consensus_pain_gi_site.csv` — ≥2 methods
- `consensus/consensus_pain_gi_site_strong.csv` — ≥3 methods
- `consensus/method_overlap_pain.csv` — per-taxon method counts

Optional: run `query_microbe_literature.py` on the strong consensus list for PubMed summaries.

# 8. Session info

```{r session}
sessionInfo()
```
